### 2022-05-31 paper.Rmd
introduction

Anticlustering is a collection of methods to partition a set of elements into disjoint groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: some anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@brusco2019; @papenberg2020; @feo1990]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis such as hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012, for an overview], anticlustering is usually formalized as a partitioning process that divides a set of elements into disjoint groups in such a way that an objective function---representing between-group similarity and/or within-group heterogeneity---is optimized [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split a test into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished by the input data and the objective function that is used to optimize group between-group similarity / within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion on the basis of pairwise dissimilarity ratings is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Maximizing the diversity optimizes within-group heterogeneity, which simultaneously leads to similar groups---if all groups have the same size. Anticlustering objective functions usually represent the reversal of a clustering objective, only the direction of the optimization is changed.  Another important criterion based on pairwise dissimilarities is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013; @brusco2019]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible.

Oftentimes, researchers work with attribute values [or *features*; @dry2009] instead of dissimilarity ratings. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing].  For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data by maximizing the k-means criterion. K-means is probably the best-known clustering method. Reversing the k-means objective function---using maximization instead of minimization---has been independently recognized as a useful anticlustering tool by Späth [-@spath1986] and Valev [-@valev1998; -@valev1983].

In k-means clustering, the sum of the squared Euclidean distances between elements and the center of the cluster they are assigned to is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. For the anticlustering application, the k-means criterion is maximized instead, thereby achieving similarity between groups. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. Hence, groups are similar with regard to the mean of the distribution of the numeric attributes wheras other parameters---such as the variance---are ignored. When requiring that groups are overall similar to each other, this focus may be misguided as similar means can be obtained even when the underlying distributions are different [e.g., @anscombe1973]. @papenberg2020 showed that maximizing the diversity (using the Euclidean distance) is better suited to minimize difference with regard to both the mean and the variance of the data; even an entirely random split usually resulted in more similar variances because k-means solely focuses on the features' means.

To optimize overall group similarity---and hence to overcome the limitations of k-means anticlustering---this paper presents k-plus anticlustering. The k-plus criterion is an extension of the k-means objective that includes one or several terms to quantify similarity with regard to other important characteristics of a data set rather than only the mean. Hence, the k-plus criterion is not a single objective function for anticlustering, but instead a familiy of objectives that extend the standard k-means objective to minimize differences between groups with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Which of these objectives is pursued depends on a researchers' needs. In the most basic---and arguably most important---case, k-plus anticlustering can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location and spread are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus objective is refered to when considering the mean and the variance of variables in the anticlustering process.

While less routinely reported, skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any other higher order moments can also be included as part of the k-plus objective. Futhermore, k-plus anticlustering can also be used to obtain similar covariance structures among groups. This approach may be of interest in cross validation applications where prediction accuracy depends on the correlations among features and criteria [@zeng2000; @papenberg2020]. Interestingly, in each case, the k-plus criterion can be reduced to an augmentation of the original data matrix, leaving the original k-means objective unchanged.

After mathematically introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups that have minimal differences according to multiple distribution characteristics. Fulfilling multiple criteria at the same time becomes more feasible as the size of the data set increases. In two pratical examples I illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

This paper focused on the presentation of new objective function but the did not really consider the multicriterion nature of the k-plus objective. Instead, a simple (unweighted) sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. Other approaches, such as direct algorithms, may yield improved results and better opportunities to select among partitionings according to multiple differently weighted k-plus criteria. @brusco2019 showed am example of how direct algorithms can be used to optimize multiple critera in anticlustering applications. Their study focused on finding the pareto set of solutions that simultaneously optimize diversity and dispersion -- convincingly arguing that applications requiring high within-group heterogeneity should incorporate both of these criteria.[^inanticlust] Future research should also investigate how k-plus anticlustering may profit from more sophisticated algorithmic treatment.

>>>>

Anticlustering is a collection of methods to partition a set of elements into disjoint groups, with the goal of obtaining high between-group similarity or within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Oftentimes, both goals coincide: some anticlustering objectives imply that similarity between groups is maximal whenever similarity within groups is minimal [@feo1990]. Anticlustering thereby reverses the logic of its better known twin cluster analysis, which seeks homogeneity within clusters and heterogeneity between clusters [@rokach2005; @steinley2006]. Whereas a variety of different approaches exist for cluster analysis---including as hierarchical methods, model-based clustering, or network approaches [see @brusco_emergent_2012, for an overview]---anticlustering is usually formalized as a partitioning process that optimizes an objective function, representing between-group similarity and/or within-group heterogeneity [@papenberg2020; @spath1986; @brusco2019].

Anticlustering is useful in many tasks surrounding research work in psychology. For example, anticlustering can be used to split tests into parts of equal difficulty [@gierl2017], to assign students to work groups [@baker2002] or when assigning stimuli to different---but parallel---experimental conditions [@lahl2006]. For a more comprehensive overview of anticlustering applications in Psychology, see @brusco2019 and @papenberg2020. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods can be distinguished on the basis of the input data and the objective function that is used to quantify between-group similarity or within-group heterogeneity. Some anticlustering objectives quantify group similarity on the basis of pairwise dissimilarity rating such as the Euclidean distance. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group. Maximizing the diversity optimizes within-group heterogeneity, which simultaneously leads to similar groups---if all groups have the same size. Another important criterion based on pairwise dissimilarities is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to find efficient solutions according to both criteria.

Oftentimes, partitioning applications in psychology use attribute data [or *features*; @dry2009] instead of dissimilarity ratings as input. This use case is for example common when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to partition the elements into groups. A different approach to anticlustering directly works with the attribute data, using an appropriate clustering objective that is computed on the basis of the features directly.  K-means is probably the best-known clustering method of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are ignored. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random split usually resulted in more similar variances. When requiring that groups are overall similar to each other, k-means' focus may be misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. The k-plus objective is an extension of the k-means objective that includes one or several terms to quantify between-group similarity with regard to other distribution characteristics and not only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance, covariances, and higher order moments such as skewness and kurtosis. Hence, the k-plus criterion is not a single criterion for anticlustering, but instead a familiy of objectives implementing different requirements with regard to between-group similarity. After mathematically introducing the k-plus objective, a simulation study is reported showing that k-plus anticlustering is well suited to create groups having minimal differences according to multiple distribution characteristics. Two pratical are included that illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust`, an extension to the popular statistical programming language R [@papenberg2020; @R-base; @R-anticlust].

@@

This paper focused on the presentation of new objective function but the did not really consider the multicriterion nature of the k-plus objective. Instead, a simple (unweighted) sum approach---combined with a generic local maximum neighbourhood search---was used to optimize the k-plus objective. Other approaches, such as direct algorithms, may yield improved results and better opportunities to select among partitionings according to multiple differently weighted k-plus criteria. @brusco2019 showed am example of how direct algorithms can be used to optimize multiple criteria in anticlustering applications. Their study focused on finding the pareto set of solutions that simultaneously optimize diversity and dispersion.[^inanticlust] Future research should also investigate how k-plus anticlustering may profit from more sophisticated algorithmic treatment.

### 2022-05-31 potpurri.Rmd
introduction

>>>>

Which of these objectives is pursued depends on a researchers' needs. In the most basic---and arguably most important---case, k-plus anticlustering can be used to not just minimize differences with regard to the mean of variables but also their variance. Measures of location and spread are arguably most important to researchers when they "eyeball" their data and they are routinely reported as characterizations of a data set. Hence, by default, the k-plus objective is refered to when considering the mean and the variance of variables in the anticlustering process.

While less routinely reported, skewness are kurtosis are also important descriptive characterizations of a data set [@decarlo1997; @westfall2014]. To optimize overall similarity between groups, it may thus be desirable to also minimize differences with regard to these distribution moments. In principle, any other higher order moments can also be included as part of the k-plus objective. Futhermore, k-plus anticlustering can also be used to obtain similar covariance structures among groups. This approach may be of interest in cross validation applications where prediction accuracy depends on the correlations among features and criteria [@zeng2000; @papenberg2020]. Interestingly, in each case, the k-plus criterion can be reduced to an augmentation of the original data matrix, leaving the original k-means objective unchanged.
